---
title: "BARBARA"
author: Peirun Chen
date: "10/25/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## K-means clustering

our first example with **kmeans()** function

```{r}
#example plot to see how R markdown works
plot(1:10,type = "l")
```

Back to kmeans
```{r}
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x=tmp, y=rev(tmp))
plot(x)
```


```{r}
#Use the kmeans() function setting k to 2 and nstart=20 # 
k <- kmeans(x, centers=2, nstart=20)
k
```

 How many points are in each cluster?
```{r}
k$cluster
```
 
 plotx colored by kmean cluster assignment and add cluster center as blue points
```{r}
palette(c("blue","green"))
plot(x,col=k$cluster)
```
 
  Q. Repeat for k=3, which has the lower tot.withinss? 
```{r}
k3 <- kmeans(x,centers = 3,nstart = 20)
k3$ tot.withinss
k$ tot.withinss
```
  
  
# Hierarchical clustering in R 

let's try out the **hclust()** function for Hierarchical clustering in R. This function need a distance matrix input

```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)
```

```{r}
plot(hc)
#Draw a line on the dendrogram
abline(h=8,col="red")

#cut the tree to tyield cluster membrane vector
cutree(hc,h=8)
```
  
  
 ### A more real example
```{r}
# Step 1. Generate some example data for clustering
x <- rbind(
 matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2), # c1
 matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2), # c2
 matrix(c(rnorm(50, mean = 1, sd = 0.3), # c3
 rnorm(50, mean = 0, sd = 0.3)), ncol = 2))
colnames(x) <- c("x", "y")
# Step 2. Plot the data without clustering
plot(x)
# Step 3. Generate colors for known clusters
# (just so we can compare to hclust results)
col <- as.factor( rep(c("c1","c2","c3"), each=50) )
palette (c ("red","blue","black"))
plot(x, col=col,pch=20)
# Q. Use the dist(), hclust(), plot() and cutree()
a <- dist(x)
hc <- hclust(a)
plot(hc)
# functions to return 2 and 3 clusters
grp.2 <- cutree(hc,h=2.5)
grp.3 <- cutree(hc,h=2)
```
Q. How does this compare to your known 'col' groups?
```{r}
table(grp.3,col)
```

 
 
 # principal component analysis

lets makes up some data with quite a few dimensions
```{r}
## Initialize a blank 100 row by 10 column matrix
mydata <- matrix(nrow=100, ncol=10)

## Lets label the rows gene1, gene2 etc. to gene100
rownames(mydata) <- paste("gene", 1:100, sep="")

## Lets label the first 5 columns wt1, wt2, wt3, wt4 and wt5
## and the last 5 ko1, ko2 etc. to ko5 (for "knock-out")
colnames(mydata) <- c( paste("wt", 1:5, sep=""),
 paste("ko", 1:5, sep="") )

## Fill in some fake read counts
for(i in 1:nrow(mydata)) {
 wt.values <- rpois(5, lambda=sample(x=10:1000, size=1))
 ko.values <- rpois(5, lambda=sample(x=10:1000, size=1))

 mydata[i,] <- c(wt.values, ko.values)
}
head(mydata)
```

Now let's try to find structure in this data with **prompt**
```{r}
pca <- prcomp(t(mydata),scale=T)
pca
```

 
  Make a PC plot of PC1 vs PC2. For this I use the "$x" component of out output (eg.pca$x)
  
```{r}
## A basic PC1 vs PC2 2-D plot
plot(pca$x[,1], pca$x[,2]) 
```
  
  Let's see how well our PC are doing
```{r}
## Precent variance is often more informative to look at
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)

pca.var.per
```


A vector of colors for wt and ko samples
```{r}
## A vector of colors for wt and ko samples
colvec <- colnames(mydata)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"

plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
 xlab=paste0("PC1 (", pca.var.per[1], "%)"),
 ylab=paste0("PC2 (", pca.var.per[2], "%)"))
```

```{r}
plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
 xlab=paste0("PC1 (", pca.var.per[1], "%)"),
 ylab=paste0("PC2 (", pca.var.per[2], "%)"))
## Click to identify which sample is which
identify(pca$x[,1], pca$x[,2], labels=colnames(mydata)) 
```

