---
title: "Hand on 9"
output: html_document
author: Peirun Chen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Preparing the data  
```{r}
wisc.df <- read.csv("https://bioboot.github.io/bimm143_S18/class-material/WisconsinCancer.csv")
```
The id and diagnosis columns will not be used for most of the following steps.
Use as.matrix() to convert the other features (i.e. columns) of the data (in columns 3 through 32) to a matrix.
```{r}
wisc.data <- as.matrix(wisc.df[3:32])
```

Set the row names of data
```{r}
rownames(wisc.data) <- wisc.df$id
```

setup a separate new vector called diagnosis to be 1 if a diagnosis is malignant ("M") and 0 otherwise.
Note that R coerces TRUE to 1 and FALSE to 0.
```{r}
table(wisc.df$diagnosis)
diagnosis <- as.numeric(wisc.df$diagnosis=="M")
sum(diagnosis)
```

How many variables/features in the data are suffixed with _mean?
```{r}
colnames(wisc.data)
s <- grep("_mean",colnames(wisc.data))
length(s)
```

#Performing PCA
check if the data need to be scaled before performing PCA
because of different units of measurement or significantly different variances.
```{r}
colMeans(wisc.data)
apply(wisc.data, 2, sd)
```

Execute PCA with the prcomp() function on the wisc.data, scaling if appropriate, and assign the output model to wisc.pr.
```{r}
wisc.pr <- prcomp(wisc.data, scale= T)
summary(wisc.pr)
```

#Interpreting PCA results
Create a biplot PC1 VS PC2. the wisc.pr using the biplot() function
```{r}
biplot(wisc.pr)
```

generate a more standard scatter plot of each observation along principal components 1 and 2
```{r}
#30 means there are 30 dimensions of PC
dim(wisc.pr$x)
#30 means there are 30 dimensions of PC
plot(wisc.pr$x[,1],wisc.pr$x[,2],col=diagnosis+1)
#because 0 has no color in palette, we add 1 so we can see 0
```

#Variance explained
```{r}
pr.var <- wisc.pr$sdev^2
sum(pr.var)
pve=pr.var/30
plot(pve,type = "o",xlab = "Principal Component",ylab = "Proportion of Variance Explained",col="blue",ylim = c(0, 1))

# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )

# Plot cumulative proportion of variance explained
plot(cumsum(pve), xlab = "Principal Component", 
     ylab = "Cumulative Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")

par(mfrow=c(1,2))
    plot(pve, xlab = "Principal Component", 
     ylab = "Cumulative Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
    plot(cumsum(pve), xlab = "Principal Component", 
     ylab = "Cumulative Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

#Hierarchical clustering of case data
Scale the wisc.data and assign the result to data.scaled.
```{r}
# Scale the wisc.data data: data.scaled
data.scaled <- scale(wisc.data)
```
we need a distance matric for hierarchical clustering input
```{r}
data.dist <- dist(data.scaled)
```
create a hierarchical clustering model using complete linkage. 
 use hclust() and assign the result to wisc.hclust
```{r}
wisc.hclust <- hclust(data.dist,method = "complete")
```

plot our tree
```{r}
plot(wisc.hclust)
```

 
 
 
 
 
 
 
 
 
 
 
 
 
#Clustering on PCA results
clustering model with complete linkage. Assign the results to wisc.pr.hclust.
```{r}
d.pr <- dist(wisc.pr$x[,1:7])
wisc.pr.hclust <- hclust(d.pr,method="complete")
plot(wisc.pr.hclust)
```

```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust,k=4)
table(wisc.pr.hclust.clusters)
```

```{r}
table(wisc.pr.hclust.clusters, diagnosis)
```







